---
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
```

```{r packages,echo=FALSE}
pacman::p_load(tidyverse,kableExtra)
```

# Freedom Data

[`Installation`](#installation) | [`Datasets`](#datasets) | [`Citations`](#citations) | [`Changelog`](#changelog)

## About

`FreedomData` is a data package that contains measures of economic freedom for 
across countries and the U.S. states from three sources: 

1. [Economic Freedom of the World (EFW)](www.fraserinstitute.org/economic-freedom/dataset) 
2. [Human Freedom Index (HFI)](https://www.cato.org/human-freedom-index/2020)
3. [Economic Freedom of North America (EFNA)](https://www.fraserinstitute.org/studies/economic-freedom-of-north-america-2020)

A link back to the GitHub repository is [here](https://github.com/acforrester/FreedomData).

## Contents

The `FreedomData` package contains the following lightly-cleaned datasets on economic and personal freedom:

1. [Economic Freedom of the World (EFW)](https://acforrester.github.io/FreedomData/reference/efw.html) 
2. [Human Freedom Index (HFI)](https://acforrester.github.io/FreedomData/reference/hfipnl.html)
3. [Economic Freedom of North America (EFNA)](https://acforrester.github.io/FreedomData/reference/efna.html)


## Installation

```{r Init, eval=FALSE}
# install.package("devtools")
devtools::install_github("acforrester/FreedomData")
```

## Datasets

### Economic Freedom of the World (EFW)

The [Economic Freedom of the World (EFW)](https://www.fraserinstitute.org/studies/economic-freedom-of-the-world-2021-annual-report) report is produced by Gwartney et al. and provides a quantitative index of the quality of countries' economic  institutions. The index contains the following elements:

| Area      | Description                       |
| :-------- | :-------------------------------- |
| Area 1    | Size of Government                |
| Area 2    | Legal System and Property Rights  |
| Area 3    | Sound Money                       |
| Area 4    | Freedom to Trade Internationally  |
| Area 5    | Regulation                        |
| Gender Legal Rights Adjustment                |

The resulting `EFW` index is an equally weighted average of each area. The index is only available for countries that have enough data to construct the index in  each year and is therefore spotty in earlier years. The number of countries in each year's `EFW` data changes as a result. Be sure to check which countries are included in your sample before conducting any empirical cross-country comparisons.

```{r efw_coverage,echo=FALSE}
# coverage by year
FreedomData::efwpnl %>% 
  group_by(`Year` = year) %>% 
  filter(!is.na(efw_index)) %>% 
  summarize(`Country Coverage` = n())%>%
  kbl(caption = "Number of EFW Countries by Year")

```

### The Human Freedom Index (HFI)

The [Human Freedom Index (HFI)](https://www.cato.org/human-freedom-index/2020) by Ian Vásquez and Fred McMahon. The HFI provides a broad-based measure of freedom for countries by combining measures of personal and economic freedom. Personal and economic freedom are measured using a combination of 76 indicators covering the areas below:

| Area             | Description                                      |
|:-----------------|:-------------------------------------------------|
| Area 1A          | Rule of Law                                      |
| Area 1B          | Security & Safety                                |
| Area 2A          | Movement                                         |
| Area 2B          | Religious Freedom                                |
| Area 2C          | Association, Assembly, &<br /> Civil Society     |
| Area 2D          | Expression & Information                         |
| Area 2E          | Identity & Relationships                         |
| Personal Freedom | Composite personal freedom index                 |    
| Economic Freedom | Economic Freedom of the World (EFW) Score        |
| Human Freedom    | Composite index of personal and economic freedom |

The `HFI` index is then an equally weighted average of personal and economic freedom. Similar to the EFW report, the number of countries covered by the HFI increases with each subsequent year.

```{r hfi_coverage,echo=FALSE}
# coverage by year
FreedomData::hfipnl %>% 
  group_by(`Year` = year) %>% 
  filter(!is.na(human_freedom)) %>% 
  summarize(`Country Coverage` = n()) %>%
  kbl(caption = "Number of HFI Countries by Year")

```

### Economic Freedom of North America (EFNA)

The [Economic Freedom of North America (EFNA)](https://www.fraserinstitute.org/studies/economic-freedom-of-north-america-2020) report from Stansel et al. provides a quantitative index that measures the quality of economic institutions at the  subnational level, viz. the U.S. states.  The index is designed to mimic the cross-country equivalent, the EFW, and is comprised of the following indicators.

| Area      | Description                       |
| :-------- | :-------------------------------- |
| Area 1    | Government Spending               |
| Area 2    | Taxes                             |
| Area 3    | Labor Market Freedom              |

The `EFNA` index itself is an equally weighted average of the three areas. The `EFNA` covers the 50 U.S. states, excluding the District of Columbia and territories.

## Citations

James Gwartney, Robert Lawson, Joshua Hall, and Ryan Murphy (2021). *Economic Freedom Dataset, published in Economic Freedom of the World: 2021 Annual Report*. Fraser Institute. [www.fraserinstitute.org/economic-freedom/dataset](www.fraserinstitute.org/economic-freedom/dataset)

Dean Stansel, José Torra, and Fred McMahon (2020). *Economic Freedom of North America 2020*. Fraser Institute. [https://www.fraserinstitute.org/studies/economic-freedom-of-north-america-2020](https://www.fraserinstitute.org/studies/economic-freedom-of-north-america-2020)

Ian Vásquez and Fred McMahon (2020). *The Human Freedom Index 2020*. Cato Institute and the Fraser Institute. [https://www.cato.org/human-freedom-index/2020](https://www.cato.org/human-freedom-index/2020)
