---
title: "An Overview of the EFNA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Overview of the EFNA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

my_colors <- function (palette = "cb") 
{
    cb_palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
        "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
    rcb_palette <- rev(cb_palette)
    bly_palette <- c("#E69F00", "#0072B2", "#000000", "#56B4E9", 
        "#009E73", "#F0E442", "#D55E00", "#CC79A7")
    
    if (palette == "cb") 
        return(cb_palette)
    else if (palette == "rcb") 
        return(rcb_palette)
    else if (palette == "bly") 
        return(bly_palette)
    else stop("Choose cb, rcb, or bly only.")
}


```

# Introduction

The *Economic Freedom of North America* (EFNA) data from Stansel et al. is a 
subnational analog to the *Economic Freedom of the World* (EFW) reports that 
measures the degree to which subnational entities in the United States, Canada, 
and Mexico support the principles of economic freedom and free exchange. The
index itself is comprised of 10 variables across three general areas listed below.

| Area      | Description                       |
| :-------- | :-------------------------------- |
| Area 1    | Government Spending               |
| Area 2    | Taxes                             |
| Area 3    | Labor Market Freedom              |

Currently, the `FreedomData` package only includes data for the U.S. states. I may 
expand this in the future to include Canada and Mexico.

# The `FreedomData` Package

In this article I highlight the usage of this simple data package and elucidate some key caveats with the EFW dataset. My baseline assumption is that the end user will make use of the common [`tidyverse`](https://tidyverse.org/) set of tools to examine and analyze these data. To highlight some features of the data of the data I will also use [`plotly`](https://plotly-r.com/) to make interactive charts.

```{r packages}
# load packages to use
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,
               plotly,
               DT)
```

## Loading the Data from the `FreedomData` Package

First we load the package.
```{r setup}
# load data
library(FreedomData)
```

I will begin with the *EFNA* panel dataset.
```{r LoadData}
data("efna")
```

## A Simple Time Series Analysis

The `EFNA` dataset covers the 50 continental U.S. states with the exclusion of 
the District of Columbia and the U.S. territories. The indexes themselves are
produced on an annual basis from 1981-present.

First, let's see what's included in the `EFNA` data by showing the top 10 most 
free states in the latest year. 

```{r header}
# view EFNA header
efna %>% 
  arrange(-year, -efna_index) %>% 
  filter(row_number() <= 10) %>% 
  modify_at(5:8, ~round(.x, 2)) %>%
  
  # subset and rename
  select(`USPS` = usps,
         `State Name` = state_name,
         `EFNA Index` = efna_index,
         `Area 1: Spending`      = efna_spending,
         `Area 2: Taxation`      = efna_taxation,
         `Area 3: Labor Markets` = efna_lab_mkt,
         ) %>% 
  knitr::kable(.)

```

Let's see how the top 5 states in 2019 evolved over time.

```{r trends}
# top and bottom
efna %>% 
  arrange(-year, -efna_index) %>% 
  filter(year == max(year),
         row_number() <= 3 | row_number() >= 48) %>% 
  pull(usps) -> top_states

p <- efna %>% 
  filter(usps %in% top_states) %>% 
  mutate(efna_index = round(efna_index, 2)) %>% 
  
  # subset and rename cols
  select(`EFNA Index` = efna_index,
         `State Name` = state_name,
         `Year`       = year) %>% 
  
  # now produce plot
  ggplot(data = .,
         aes(x = `Year`,
             y = `EFNA Index`,
             color = `State Name`)) +
  geom_line() +
  theme_minimal() +
  scale_color_manual(values = my_colors("cb")[1:6],
                    guide = "none")+
  labs(title = "Economic Freedom by State, 1981-2018",
       caption = "Source: Stansel et al. (2020)",
       x = "",
       y = "Avg. Economic Freedom Score")

# render plot
plotly::ggplotly(p)

```

# References

Dean Stansel, José Torra, and Fred McMahon (2020). *Economic Freedom of North America 2020*. Fraser Institute. [https://www.fraserinstitute.org/studies/economic-freedom-of-north-america-2020](https://www.fraserinstitute.org/studies/economic-freedom-of-north-america-2020)
